---
title: "Project_1"
author: "Juliette Licon jtl2383"
date: "3/15/2020"
output:
  pdf_document: default
  html_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

```{r}
library(plyr)
library(tidyverse)
library(plotly)
library(cluster)
library(psych)


GradRates17<-read.csv("State_Grad_Rates_2017.csv")
Poverty17<-read.csv("State_Poverty_2017.csv")
Budget17<-read.csv("State_Budget_2017.csv")
```

*For the year of 2017, I have compiled 3 different datasets, comparing each of the states in the US. The first, entitled GradRates17, encompasses the graduation rate, Poverty17 holds the number of poverty cases and the total population of each state, and Budget17 contains the total budget, budget towards mental health care, and budget towards education. While the information was published online, I input the data into a csv file manually. State budgets were acquired from Ballotpedia.org as well as populations. Graduation rates were acquired from Eduweek.org, and poverty rates were acquired from www.census.gov.*

*These datasets were particularly interesting to me as this begins to delve into what I will be researching in my master's program for Public Health. I'm very interested in how mental health care, provided through the means of public education systems, affects children and their futures, hence the graduation and poverty rates in my set. While this project is only a small snippet into potential associations between the prioritization of a state's health care and education, it could provide insight into which states could potentially be used as models for states with higher poverty and lower graduation rates. As of now, I would expect to see a positive correlation between education and mental health spending and graduation rates, and a a negative association with poverty rates.*


*The code below is transforming each separate dataset into data frames that will seamlessly combine. In GradRates17, pivot_wider() widened the variables 'State' and 'Graduation.Rates.Percent' into a dataframe where each state is a separate variable with a value of the graduation rate. pivot_longer() recreates the table containing the two variables "State' and 'Graduation',containing the graduation rate. as.factor() transforms the State variable from a character variable into a factor variable for GradRates17, Budget17, and Poverty17. mutate() creates a new variable with identical data from 'Total.Spending', but labeled as 'Total Budget'. An additional pivot_longer() takes all the variables, aside from 'State' and 'Total Budget', and creates a new categorical variable, 'Budget.Type', clarifying the type of budget, either Education, Mental Health Care, or Total, and a new numerical variable of how much money was spent on each budget.*

```{R}
GradRates17<-GradRates17%>%pivot_wider(names_from="State",values_from="Graduation.Rate.Percent")%>%pivot_longer(1:50,names_to="State", values_to="Graduation")
GradRates17$State<-as.factor(GradRates17$State)


Poverty17$State<-as.factor(Poverty17$State)


Budget17<-Budget17%>%mutate(`Total Budget`=Total.Spending)%>%pivot_longer(2:4, names_to="Budget.Type", values_to="Amount.Spent")
Budget17$State<-as.factor(Budget17$State)
```

*Here the 3 datasets were all joined by matching on the variable "State. Since there were no NAs in any of the datasets, and each contained all 50 states, full_join() was used to ensure combination and retention of all values and variables. In the 'FullData' dataset, the new variables were 'State'(factor), 'Graduation'(double class due to the decimal), 'Poverty.Cases'(integer), 'Total.Population'(integer), 'Total Budget'(double class), 'Budget.Type'(character), and 'Amount Spent' per budget(double class). The first few lines rows of data can be viewed by glimpse().*
```{R}
FullData<-GradRates17%>%full_join(Poverty17, by="State")%>%full_join(Budget17, by="State")%>%glimpse()
```

*Two new datasets were create: EduData and MentalData. EduData was created by selecting only the rows from FullData that contained 'Education.Spending' as the Budget Type. Then a new variable, 'Education Percentage of Total Budget(%)' was created by dividing the amount spent on the education budget by the total budget and multiplying the result by 100. A new variable, 'Poverty Rate(%)' was created by dividing the column of poverty cases by total populations for each state and multiplying by 100. Then only the rows of State, graduation rates, poverty rates, and education percentage of total budgets were kept. The same process was repeated for MentalData, but only rows from FullData that contained 'Mental.Health.Spending' as the budget type,and the variable 'Mental Health Percentage of Total Budget(%)' was created and kept. FullData1 was the full join of EduData and MentalData, which combined on the variables 'State', 'Graduation' and 'Poverty Rate(%)' and 'Graduation' was renamed as 'Graduation Rate(%)'. Then the variables were reordered in the data frame to have State first, followed by Poverty Rate, Graduation rate, education percentage of total budget, and mental health percentage of total budget.*
```{r}
EduData<-FullData%>%filter(`Budget.Type` == "Education.Spending")%>%mutate(`Education Percentage of Total Budget(%)` = Amount.Spent/`Total Budget`*100, `Poverty Rate(%)`=Poverty.Cases/Total.Population*100)%>%select(1,2,8:9)

MentalData<-FullData%>%filter(`Budget.Type` == "Mental.Health.Spending")%>%mutate(`Mental Health Percentage of Total Budget(%)` = Amount.Spent/`Total Budget`*100, `Poverty Rate(%)`=Poverty.Cases/Total.Population*100)%>%select(1,2,8:9)

FullData1<-EduData%>%full_join(MentalData)%>%select(State,`Poverty Rate(%)`, `Graduation Rate(%)`="Graduation", everything())%>%glimpse
```

*To create a new categorical variable describing the priority a state placed mental health care and education budgetwise, a blank variable, 'Spending Priority' was first created and filled with NAs. Then this was filled depending on which logical condition the state fulfilled. A state could either be high or low priority on education depending on if they spent higher or lower than 15% of their budget on education, and high or low on mental health priority depending on if they spent higher or lower than 2% of their budget on mental health care. While I set the threshholds myself, for a more in depth project, I would use more research to determine the cutoffs.*
```{r}
FullData1$`Spending Priority`<-NA
FullData1[FullData1$`Education Percentage of Total Budget(%)`>15 & FullData1$`Mental Health Percentage of Total Budget(%)`>2, "Spending Priority"]<-"High Education & Mental Health"

FullData1[FullData1$`Education Percentage of Total Budget(%)`>15 & FullData1$`Mental Health Percentage of Total Budget(%)`<2, "Spending Priority"]<-"High Education Only"

FullData1[FullData1$`Education Percentage of Total Budget(%)`<15 & FullData1$`Mental Health Percentage of Total Budget(%)`>2, "Spending Priority"]<-"High Mental Health Only"

FullData1[FullData1$`Education Percentage of Total Budget(%)`<15 & FullData1$`Mental Health Percentage of Total Budget(%)`<2, "Spending Priority"]<-"Low Education & Mental Health"

FullData1%>%glimpse()
```

*For each pair of coding lines, the first line is a summary statistic for each variable utilizing the entire dataset, while the second line is calculating the summary statistic after grouping by the Spending Priority.The statistics calculated for each are mean, standard deviation, minimum value, maximum value, and the variance.*
```{r}
FullData1%>%summarize_if(is.numeric, mean)
FullData1%>%group_by(`Spending Priority`)%>%summarize_if(is.numeric, mean)

FullData1%>%summarize_if(is.numeric, sd)%>%print()
FullData1%>%group_by(`Spending Priority`)%>%summarize_if(is.numeric, sd)%>%print()

FullData1%>%summarize_if(is.numeric, min)%>%print()
FullData1%>%group_by(`Spending Priority`)%>%summarize_if(is.numeric, min)%>%print()

FullData1%>%summarize_if(is.numeric, max)%>%print()
FullData1%>%group_by(`Spending Priority`)%>%summarize_if(is.numeric, max)%>%print()

FullData1%>%summarize_if(is.numeric, var)%>%print()
FullData1%>%group_by(`Spending Priority`)%>%summarize_if(is.numeric, var)%>%print()
```

*These 2 lines of code create 2 tables of all summary statistics, including the quartiles. The second table produces the summary statistics grouped by Spending Priority.*
```{r}
FullData1 %>%select_if(is.numeric)%>% describe(quant = c(.25,0.5, .75)) %>% print()
FullData1%>%select_if(is.numeric)%>%describe.by(FullData1$`Spending Priority`, quant=c(0.25,0.5,0.75))%>%print()
```

*FullData2 dataset was created by only taking the numeric variables from the dataset of FullData1 (everything besides State), scaling it to z-scores, and converting it to a data frame. Then a correlation matrix was created and correlation values rounded to 2 decimal points.*
```{r}
FullData2<-FullData1%>%select_if(is.numeric)%>%scale()%>%data.frame()

FullData2%>%cor()%>%round(2)
```

*Using FullData1, numeric variables were selected and run through a correlation matrix and coereced into a data frame that named the rows by transposing the columns names, then the data frame was pivoted longer to create rows that consisted of each combination of the numeric variables with the third variable being the correlation value. Using ggplot and geom_tile(), a 4 by 4 tile matrix was created matching up every combination of variables, and filling in the interesecting tiles with their correlation value. The darker purple a square is, the higher positive correlation exists and the darker green a square is, the higher negative correlation exists between the two variables. The titles of the horizontal axis were then titled to be at a 45 degree angle so as it make it easier to read without any overlap of the titles.The highest correlation that existed was a negative correlation between poverty rates and graduation rates, followed by an equal negative correlation between education percentage of budget and poverty rate, and mental health percentage of budget and poverty rate.The shocking negative correlation was the negative correlation between mental health percentage and graduation rates.*
```{r}
FullData1%>%select_if(is.numeric)%>%cor%>%as.data.frame%>%rownames_to_column%>%pivot_longer(-1)%>%ggplot(aes(rowname,name,fill=value))+geom_tile()+geom_text(aes(label=round(value,2)))+xlab("")+ylab("")+scale_fill_gradient2(low="green",high="purple")+theme(axis.text.x = element_text(angle=45, hjust=1))
```

*A scatterplot was created by mapping graduation rate and poverty rate as the axes, where we expect to see a negative correlation. The color of each point was determined by the state's spending priority. While there is no obvious clustering between the different spending priorities, we can see that the states with the highest poverty rates and lowest graduation rates were all states that had low priority spending on education AND mental health care.*
```{r}
ggplot(FullData1,aes(x=`Graduation Rate(%)`,y=`Poverty Rate(%)`, color=`Spending Priority`))+geom_point()
```

*This bar graph shows the states ordered by graduation rate and  colored by spending priority. The coordinates were flipped to create a horizontal bar graph. Since there was only one value for each state, no mean or standard error could be applied. There was no obvious patterns available, but it was shocking that states such as Michigan, Washington, and Arizona, were considered high for priority of education and mental health, yet yielded some of the lowest graduation rates. While they spent a larger portion of their budget on these sectors, their per capita expenditure may actually be much lower than states with higher graduation rates.*
```{r}
ggplot(FullData1, aes(x =reorder(State,-`Graduation Rate(%)`), y = `Graduation Rate(%)`, fill=`Spending Priority`))+
  geom_bar(stat="summary",width=0.5)+
  theme(axis.text.x = element_text(angle=90, hjust=1))+coord_flip()+xlab("State")
```

*PAM clustering was utilized first creating a pam_dat dataset composed of all 4 numeric variables atfer scaling their values. To determine how many clusters we should use, average silhoutte width needs to be calculated. First an empty vector was created to hold the average silhouette widths. Then the for loop was used to compute the PAM values, then take the average silhouette widths. A line plot was created out of the average silhoutte widths and 2 clusters yielded the highest average, making it the best number of clusters to use. However, we want to see if we can cluster the states into their spending priorities, which would have us use 4 clusters, and 4 clusters did acually yield the second largest average silhouette width.*
```{r}
pam_dat<-FullData1%>%select(`Mental Health Percentage of Total Budget(%)`,`Education Percentage of Total Budget(%)`,`Poverty Rate(%)`,`Graduation Rate(%)`)%>%scale()
sil_width<-vector()
for(i in 2:10){
pam_fit <- pam(pam_dat, k = i)
sil_width[i] <- pam_fit$silinfo$avg.width
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)

```

*pam1 was then created from FullData, selecting all the numeric variables and scaling them, and running them through pam clustering with 2 clusters. The two medoids found were State 33, 2, 7, and 47, which correlate to North Carolina, Alaska, Connecticut, and Washington, respectively. PLotting the acerage silhouette widths give us an avergae silhouette width of 0.24 which means no substantial structure has been found. Even with 2 clusters instead of 4, the average silhouette width is only 0.26, just barely a weak structure that could be artificial.*
```{r}
pam1<-FullData1%>%select_if(is.numeric)%>%scale()%>%pam(4)
pam1
pam2<-FullData2%>%select_if(is.numeric)%>%scale()%>%pam(2)
plot(pam1, which=2)
plot(pam2, which = 2)
```

*To determine if the clusters were grouped well we will add a new variable to the FullData1 with factor values of which cluster each state was put in. Then a new dataset, confmat was created by grouping the States by their Spending Priority and counting up how many from each spending priority made it into each cluster. The data was then rearranged to make the rows the spending priorites and the columns the clusters with the counts as values. Then to determine the accuracy of the clustering, confmat was coerced into a matrix and then determining the hit rate. The hit rate is determined by counting how many of each spending priority group made it into the separate diagonal portion of the matrix. If all the states were clustered perfectly, then all of one spending priority would be in cluster 1, all of another spending priority would be in cluster 2 and so on. We produced a hit rate of 36% accuracy.* 
```{r}
final<-FullData1%>%mutate(cluster=as.factor(pam1$clustering))

confmat<-final%>%group_by(`Spending Priority`)%>%count(cluster)%>%arrange(desc(n))%>%pivot_wider(names_from="cluster",values_from="n",values_fill = list('n'=0))
confmat

round(sum(diag(as.matrix(confmat[,2:5])))/sum(confmat[,2:5]),5)
```

*Using Graduation and Poverty rates as the two axes to plot each state on, due to their highest correlation values, each state was colored by cluster. Cluster 1 overlaps greatly with both 2 and 3, but 2 and 3 are separated fairly well. Cluster 4 only contains one point in the middle of all the clusters.*
```{r}
ggplot(final, aes(x=`Graduation Rate(%)`,y=`Poverty Rate(%)`, color=cluster))+geom_point()
```

*A 3D plot was then made by adding in the variable mental health percentage of the total budget and spending priority was shown by shape of each point. While clusters were fairly well grouped, states of each each spending priority were scattered amongst the different clusters.*
```{r}
final%>%plot_ly(x= ~`Graduation Rate(%)`, y = ~`Poverty Rate(%)`, z = ~`Mental Health Percentage of Total Budget(%)`, color= ~cluster,type = "scatter3d", mode = "markers", symbol = ~`Spending Priority`, symbols = c('circle','x','square'))
```

```{R}